NDNS multilevel LCA CW2CB2 or CW2CB3

### 2018-08-20 paper/Chapters/Chapter4.tex
abstract almost done, proof reading first round done

Surprisingly, low carbohydrate eaters whose dietary recordings suggested that they were mostly following a regular temporal meals pattern turned out to be consuming the highest amount of total energy among the three types of carbohydrate eaters. Detailed profiling of energy composition according to the time slots revealed that low carbohydrate eaters actually had higher proportion of energy contributed by both alcohol and fat. High percentage of fat consumption was shown in all 7 time slots, energy coming from alcohol exceeded more than one fifth of the total energy after 8 pm. These findings explained why they were actually consuming the highest energy among the three types of carbohydrate eaters. However, we also found that participants consuming low carbohydrate food had higher prevalence of diabetes, hypertension, and obesity. These health issues might possibly lead them (or advices were given from their physicians) to replace carbohydrates in their diet to other energy sources such as fat, protein, or even alcohol. Therefore, there is a possibility that they chose to follow low carbohydrate diets out of health purpose, but many of them were replacing carbohydrates with higher energy condensed food or even alcohol at night which might indeed be a public health concern.

@@

Among men, who were classified as moderate carbohydrate eaters probably had lower odds of having hypertension, after adjustment of age, live with partner or not, educational level, BMI, smoking status and total energy intake. As discussed above, moderate carbohydrate eaters turned to have meals (or energy consumption) later in time compared with high carbohydrate eaters, but low carbohydrate eaters also consumed large amount of energy (from both fat and alcohol) at night. Therefore, considering that moderate carbohydrate eaters were younger than low carbohydrate eaters (although age was adjusted in the full models), there is probably reverse causality exists here (also due to the nature of cross-sectional study). That is, they were potentially both late eaters, however, with increased age (and so as increased health-related problems/concern) some of them modified their habits, such as quit smoking, replace carbohydrate food with other energy sources which lead to the phenomenon of lower odds of hypertension in moderate carbohydrate eaters. Although these hypotheses cannot be determined by the cross-sectional data from NDNS RP, if they were true, the energy sources they used to replace carbohydrate in their diet were apparently not very wisely chosen.

Among women, whether living with a partner became an interaction factor for the associations between carbohydrate eating patterns and BMI and abdominal obesity (WC). Directions of the associations were opposite to each other depending on whether women were living with a partner or not. This interaction effect was more obvious when looking at abdominal obesity measurement. High carbohydrate eaters had lower BMI and WC in those lived with their partner, while moderate carbohydrate eaters had higher WC in those who lived alone after adjustment of age, education level, smoking status, total energy intake, and alcohol consumption. High carbohydrate eaters who were characterised with high and early in the day energy consumption and low fat and alcohol intake may reflect a healthier diet and lifestyle, but this might be different between women who lived alone and those who lived with their partners. It was often assumed that live alone may associated with lower diversity of food intake, and a higher likelihood of having an unhealthy dietary pattern \parencite{hanna2015relationship}. Therefore, there may be differences in the actual contents consumed in high carbohydrates eating women, or there may be other social, psychological or lifestyle related factors related with living alone which were not measured or did not included in the models, so that the inverse association between high carbohydrate eating pattern and BMI or abdominal obesity were only observed among women who lived with partners, further analysis is needed. Whereas the reason why moderate carbohydrate eaters' WC was larger than low carbohydrate eaters only among women who lived alone is unknown, given that the evidence of this association was weak and borderline significant, whether it was just a false positive finding should be explored in other studies.

@@

Firstly, we ignored the order of observation days in the MLCA models. The food consumption diaries were accomplished by participants for at least 3 out of 4 consecutive days. In the multilevel analyses, these 3 or 4 days' observations were treated as if they were exchangeable in the models. Since one's diet might change according to season, day of the week (weekdays or weekends), or sometimes depends on what they had consumed the day before, exchangeability of their diaries is a strong assumption and cannot be overcome by the MLCA models adopted here. Other statistical techniques, which could take the order or the longitudinal nature of the data into account, such as repeated measures latent class analysis (RMLCA), latent transition analysis (LTA) \parencite{collins2010latent}, or latent class growth analysis (LCGA) \parencite{davidian2008growth,jung2008introduction,andruff2009latent} are not applicable for the NDNS RP dataset. For the purpose of maintaining the response rate, flexibility was allowed for the NDNS RP participants in their choice of which day to begin their diary. But the above mentioned alternative models require that the 3 or 4 repeated measurements of food diaries to be recorded at the same time points longitudinally. Specifically, LCGA models (also called growth mixture modelling) will also need to model the change of the odds of the probability over time as a function (quadratic, or cubic) which is apparently not the objective in the current project.

Secondly, the classification of individuals to latent carbohydrate eating classes were defined by maximum posterior probability assignment rule. This approach assigns individuals to the class for which they have the highest posterior probability of membership \parencite{nagin2005group}. The other approach is multiple pseudo-class draws \parencite{wang2005residual}, which was proposed in attempt to account for uncertainty in class assignment. However, the maximum probability rule is still believed to be able to minimize the number of incorrect assignments \parencite{goodman20071}. A Monte Carlo simulation study \parencite{bray2015eliminating} also demonstrated that maximum probability assignment is less biased than multiple pseudo-class draws. Moreover, this simulation study also found that an inclusive LCA (i.e. LCA with covariates) would perform better than non-inclusive LCA, and has the potential to reduce bias of class assignment. However, whether this advantage can be extended to MLCA in the current project is unknown, and also due to the reason that the associations between the latent classes and distal outcomes (hypertension, and BMI) in our analyses are mostly still under exploration, therefore, the model for LCA with covariates can never be well-known before our analyses. Thus, as a first step, given the complexity of the NDNS RP dataset itself, we chose to conduct the MLCA modellings without any covariate in either day level or individual level models. Future studies may be advised to consider incorporating other predictors in the MLCA models to see whether the classifications in both levels can be improved or not.

Thirdly, information regarding the detailed occupations of the participants was not available and because of which we could not exclude those who had a shift work involving night work. But considering that completing a 4-consecutive-day food diary would be a large burden for the participants, those had a shift work involving night work might possibly performed relatively lower compliance and could not complete their diary for at least 3 days. But this hypothesis cannot be verified in the NDNS RP dataset. If people working with shift work were included in the NDNS RP sample, the classification in both levels as well as their associations with either hypertension or BMI would be biased. Furthermore, under-reporting due to the burden of writing down everything they eat at each occasion is also a concern in the food diaries collected. To evaluate the influence of under-reporting in the NDNS RP sample, study team of NDNS RP conducted doubly labelled water (DLW) sub-study within the survey. The details of this sub-study can be accessed from the Appendix X of the Official Reports provided by Public Health England \parencite{bates2014national,roberts2018national,NDNSofficial}. Briefly, in healthy adults, if, for a given period of time, energy consumed in food matches total energy expended, they are in energy balance. In NDNS RP sub-study, estimates of energy intake (EI) from the four-day diary were compared with measurements of total energy expenditure (TEE) using the DLW technique in a sub-sample of survey participants. Results of the sub-study showed reasonable agreement between EI and TEE (overall ratio: 0.73). Reasons such as misreporting of actual consumption; under-reported or modified usual intake due to the burden of the survey/DLW sub-study are potential factors may contributed to under-reporting of their EI. We cannot extrapolate the estimated under-reporting to the whole sample since other individuals' diet might be affected by under-reporting differentially.

Lastly, it should be noted that findings from data-driven, exploratory methods may not be
generalizable to populations in other countries. Since the carbohydrate eating patterns in both day level and individual level probably only represent the socio-cultural and life-style characteristics of the adult population in the UK. Further researches are warranted to explore and better understand the eating patterns in other populations. Besides, the cross-sectional study design also refrained us from deducing any causal effect between the carbohydrate eating patterns and hypertension or BMI.

Strengths of this study include the large, nationally representative sample of men and women in the UK. We applied a novel, objective approach using MLCA to examine the carbohydrate eating patterns while using standardised criteria to determine the numbers of latent classes. The process of finding the classifications through model-based, data-driven procedure minimises reliance on researchers' preconcieved notions of eating patterns. Eating patterns were determined from 4 consecutive days of well-designed, fully-examined food diaries. MLCA correctly accounted for the multilevel structure of the data which means that the 4-day diet diaries were nested within the participants. Our findings also captured the day-to-day variation of following different carbohydrate eating patterns within individuals. Moreover, the examination of associations between the individual level carbohydrate eating patterns and hypertension, BMI, abdominal obesity were conducted correctly considering the probability of participant selection and the clustered survey design. Health outcomes such as weight, height, waist circumferences, and blood pressure, diabetes status were measured by trained nurses and blood tests which could also minimise the misclassification and under-reporting of these measurements.

@@

We have successfully defined carbohydrate eating patterns in the general population in the UK adults using the NDNS RP database in both observation day level and individual level. Low carbohydrate eaters turned to have more energy that contributed by both fat and alcohol. Moderate carbohydrate eaters consumed the lowest total energy, while they had the tendency of having meals later in time-of-day. High carbohydrate eaters consumed most of their carbohydrate as well as energy earlier in time-of-day. These dietary patterns specifically for carbohydrate intake were found to be differed by timing, quantity and contributions to energy consumption. Compared with low carbohydrate eaters, men had moderate carbohydrate eating pattern may associated with a lower prevalence of hypertension, women in this latent class who lived alone may associated with a larger waist circumferences. Among women who lived with partners, high carbohydrate eating pattern was associated with both lower BMI and smaller waist circumferences. Longitudinal designed studies are needed in finding whether the eating patterns themselves are changing over time and in investigating how such circadian eating patterns may relate to the change of blood pressure, obesity, and even mortality.

>>>>

Surprisingly, low carbohydrate eaters whose dietary recordings suggested that they were mostly following a regular temporal meals pattern turned out to be consuming the highest amount of total energy among the three types of carbohydrate eaters. Detailed profiling of energy composition according to the time slots revealed that low carbohydrate eaters actually had higher proportion of energy contributed by both alcohol and fat. High percentage of fat consumption was shown in all 7 time slots, energy coming from alcohol exceeded more than one fifth of the total energy after 8 pm. These findings explained why they were actually consuming the highest energy among the three types of carbohydrate eaters. In the meantime, we also found that participants consuming low carbohydrate food had higher prevalence of diabetes, hypertension, and obesity. These health issues might possibly lead them (or advices were given from their physicians) to replace carbohydrates in their diet to other energy sources such as fat, protein, or even alcohol. Therefore, there is a possibility that they chose to follow low carbohydrate diets out of health purpose, but many of them were replacing carbohydrates with higher energy condensed food or even alcohol at night which might indeed be a public health concern.

@@

Among men, who were classified as moderate carbohydrate eaters probably had lower odds of having hypertension, after adjustment of age, live with partner or not, educational level, BMI, smoking status and total energy intake. As discussed above, moderate carbohydrate eaters turned to have meals (or energy consumption) later in time compared with high carbohydrate eaters. But, it is noteworthy that low carbohydrate eaters also consumed large amount of energy (from both fat and alcohol) at night. Therefore, considering that moderate carbohydrate eaters were younger than low carbohydrate eaters (although age was adjusted in the full models), there is probably reverse causality exists here (also due to the nature of cross-sectional study). That is, they were potentially both late eaters, however, with increased age (and so as increased health-related problems/concern) some of them modified their habits, such as quitting smoking, replacing carbohydrate food with other energy sources (so that they became low carbohydrate eaters with hypertension) which lead to the phenomenon of lower odds of hypertension in moderate carbohydrate eaters. Although these hypotheses cannot be determined by the cross-sectional data from NDNS RP, if they were true, the energy sources they used to replace carbohydrate in their diet were apparently not very wisely chosen.

Among women, whether living with a partner became an interaction factor for the associations between carbohydrate eating patterns and BMI and abdominal obesity (WC). Directions of the associations were opposite to each other depending on whether women were living with a partner or not. This interaction effect was more obvious when looking at abdominal obesity measurement. High carbohydrate eaters had lower BMI and WC in those lived with their partners, while moderate carbohydrate eaters had higher WC in those who lived alone after adjustment of age, education level, smoking status, total energy intake, and alcohol consumption. High carbohydrate eaters who were characterised with high and early in the day energy consumption and low fat and alcohol intake may reflect a healthier diet and lifestyle, but this might be different between women who lived alone and those who lived with their partners. It was often assumed that live alone may associated with lower diversity of food intake, and a higher likelihood of having an unhealthy dietary choice \parencite{hanna2015relationship}. Therefore, there may be differences in the actual contents consumed in their high carbohydrates, or there may be other social, psychological or lifestyle related factors related with living alone which were not measured or did not included in the models. Thus, the inverse association between high carbohydrate eating pattern and BMI or abdominal obesity were only observed among women who lived with partners, further investigation on this hypothesis is needed. Whereas the reason why moderate carbohydrate eaters' WC was larger than low carbohydrate eaters only among women who lived alone is unknown, given that the evidence of this association was weak and borderline significant, whether it was just a false positive finding should also be explored in other studies.

@@

Firstly, we ignored the order of observation days in the MLCA models. The food consumption diaries were accomplished by participants for at least 3 out of 4 consecutive days. In the multilevel analyses, these 3 or 4 days' observations were treated as if they were exchangeable in the models. Since one's diet might change according to season, day of the week (weekdays or weekends), or sometimes depends on what they had consumed the day before, exchangeability of their diaries is a strong assumption and cannot be overcome by the MLCA models adopted here. Other statistical techniques, which could take the order or the longitudinal nature of the data into account, such as repeated measures latent class analysis (RMLCA), latent transition analysis (LTA) \parencite{collins2010latent}, or latent class growth analysis (LCGA) \parencite{davidian2008growth,jung2008introduction,andruff2009latent} are not applicable for the NDNS RP dataset. For the purpose of maintaining the response rate, flexibility was allowed for the NDNS RP participants in their choice of which day to begin their diary. But the above mentioned alternative models require that the 3 or 4 repeated measurements of food diaries to be recorded at the same time points longitudinally. Specifically, LCGA models (also called growth mixture model) will also need to model the change of the odds of the probability over time as a function (quadratic, or cubic) which is apparently not the objective in the current project.

Secondly, the classification of individuals to latent carbohydrate eating classes were defined by maximum posterior probability assignment rule. This approach assigns individuals to the class for which they have the highest posterior probability of membership \parencite{nagin2005group}. The other approach is multiple pseudo-class draws \parencite{wang2005residual}, which was proposed in attempt to account for uncertainty in class assignment. However, the maximum probability rule is still believed to be able to minimize the number of incorrect assignments \parencite{goodman20071}. A Monte Carlo simulation study \parencite{bray2015eliminating} demonstrated that maximum probability assignment is less biased than multiple pseudo-class draws. Moreover, this simulation study also found that an inclusive LCA (i.e. LCA with covariates) would probably perform better than non-inclusive LCA, and has the potential to reduce bias of class assignment. However, whether this advantage can be extended to MLCA in the current project is unknown, and also due to the reason that the associations between the latent classes and distal outcomes (hypertension, and BMI) in our analyses are mostly still under exploration, therefore, the model for LCA with covariates can never be well-known before our analyses. Thus, as a first step, given the complexity of the NDNS RP dataset itself, we chose to fit the MLCA models without any covariate in either day level or individual level models. Future studies may be advised to consider incorporating other predictors in the MLCA models to see whether the classifications in both levels can be improved or not.

Thirdly, information regarding the detailed occupations of the participants was not available and because of which we could not exclude those who had a shift work involving night work. But considering that completing a 4-consecutive-day food diary would be a large burden for the participants, those had a shift work involving night work might possibly performed with relatively lower compliance and could not complete their diary for at least 3 days. But this hypothesis cannot be verified in the NDNS RP dataset. If people working with shift work were included in the NDNS RP sample, the classification in both levels as well as their associations with either hypertension or BMI would be biased. Furthermore, under-reporting due to the burden of writing down everything they eat at each occasion is also a concern in the food diaries collected. To evaluate the influence of under-reporting in the NDNS RP sample, study team of NDNS RP conducted doubly labelled water (DLW) sub-study within the survey. The details of this sub-study can be accessed from the Appendix X of the Official Reports provided by Public Health England \parencite{bates2014national,roberts2018national,NDNSofficial}. Briefly, in healthy adults, if, for a given period of time, energy consumed in food matches total energy expended, they are in energy balance. In NDNS RP sub-study, estimates of energy intake (EI) from the four-day diary were compared with measurements of total energy expenditure (TEE) using the DLW technique in a sub-sample of survey participants. Results of the sub-study showed reasonable agreement between EI and TEE (overall ratio: 0.73). Reasons such as misreporting of actual consumption; under-reported or modified usual intake due to the burden of the survey/DLW sub-study are potential factors may contributed to under-reporting of their EI. We cannot extrapolate the estimated under-reporting to the whole sample since other individuals' diet might be affected by under-reporting differentially.

Lastly, it should be noted that findings from data-driven, exploratory methods may not be
generalizable to populations in other countries. Since the carbohydrate eating patterns in both day level and individual level probably only represent the socio-cultural and life-style characteristics of the adult population in the UK. Further researches are warranted to explore and better understand the eating patterns in other populations. Besides, the cross-sectional study design also refrained us from deducing any causal effect between the carbohydrate eating patterns and hypertension, BMI or abdominal obesity.

Strengths of this study include the large, nationally representative sample of men and women in the UK. We applied a novel, objective approach using MLCA to examine the carbohydrate eating patterns while applying standardised criteria to determine the numbers of latent classes. The process of finding the classifications through model-based, data-driven procedure minimises reliance on researchers' preconcieved notions of eating patterns. Eating patterns were determined from 4 consecutive days of well-designed, fully-examined food diaries. MLCA correctly accounted for the multilevel structure of the data which means that the 4-day diet diaries were nested within the participants. Our findings also captured the day-to-day variation of following different carbohydrate eating patterns within individuals. Moreover, the examination of associations between the individual level carbohydrate eating patterns and hypertension, BMI, abdominal obesity were conducted correctly considering the probability of participant selection and the clustered survey design. Health outcomes such as weight, height, waist circumferences, and blood pressure, diabetes status were measured by trained nurses and blood tests which could also minimise bias caused by misclassification and under-reporting of these measurements.

@@

We have successfully defined carbohydrate eating patterns in the general population in the UK adults using the NDNS RP database in both observation day level and individual level. Low carbohydrate eaters turned to have more energy that contributed by both fat and alcohol. Moderate carbohydrate eaters consumed the lowest total energy, while they had the tendency of having meals later in time-of-day. High carbohydrate eaters consumed most of their carbohydrate as well as energy earlier in time-of-day. These dietary patterns specifically for carbohydrate intake were found to be differed by timing, quantity and contributions to energy consumption. Compared with low carbohydrate eaters, men had moderate carbohydrate eating pattern may associated with a lower prevalence of hypertension, women in this latent class who lived alone may associated with a larger waist circumferences. Among women who lived with partners, high carbohydrate eating pattern was associated with both lower BMI and smaller waist circumferences. Longitudinal designed studies are needed in finding whether the eating patterns themselves are changing over time and in investigating how such circadian eating patterns may relate to the change of blood pressure, obesity, and other health outcomes (incidence of cancer, cardiovascular diseases, or perhaps mortality).

### 2018-08-20 paper/main.tex
abstract almost done, proof reading first round done

\subject{Medical Statistics} % Your subject area, this is not currently used anywhere in the template, print it elsewhere with \subjectname
\keywords{} % Keywords for your thesis, this is not currently used anywhere in the template, print it elsewhere with \keywordnames
\university{\href{https://www.lshtm.ac.uk/}{London School of Hygiene and Tropical Medicine}} % Your university's name and URL, this is used in the title page and abstract, print it elsewhere with \univname
\department{\href{}{Page count: 42 from Introduction to Conclusions}} % Your department's name and URL, this is used in the title page and abstract, print it elsewhere with \deptname
\group{Candidate number: 110765} % Your research group's name and URL, this is used in the title page, print it elsewhere with \groupname
\faculty{\href{http://faculty.university.com}{Faculty Name}} % Your faculty's name and URL, this is used in the title page and abstract, print it elsewhere with \facname

@@

\begin{declaration}
\addchaptertocentry{\authorshipname} % Add the declaration to the table of contents
\noindent I, \authorname, declare that this thesis titled, \enquote{\ttitle} and the work presented in it are my own. I confirm that:

@@

\noindent Signed: \;\;\;\;\;\;\includegraphics[height=2.5\baselineskip]{Figures/signature.png}\\
\rule[0.5em]{25em}{0.5pt} % This prints a line for the signature

\noindent Date: \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;2018-09-05\\
\rule[0.5em]{25em}{0.5pt} % This prints a line to write the date
\end{declaration}

@@

\begin{acknowledgements}
\addchaptertocentry{\acknowledgementname} % Add the acknowledgements to the table of contents
\vspace{0.8cm}
I would like to thank my tutor and supervisor \supname, for his guidance, patience, and help while working on this project and also to Dr. Suzana Almoosawi for her invaluable nutritional academic insight, and suggestions. \\

@@

\begin{abstract}
\addchaptertocentry{\abstractname} % Add the abstract to the table of contents
The National Diet and Nutrition Survey (NDNS) database of detailed four-day food diaries was used in exploration of both temporal and quantity eating patterns of carbohydrates in the general adults living in the UK. 6155 participants aged 19 or older were kept in the analysis. \\ 
Multilevel latent class analysis (MLCA) models were applied in finding the latent classes of carbohydrate consumption both at the observation day level, and individual level.
\end{abstract}

>>>>

\subject{Medical Statistics} % Your subject area, this is not currently used anywhere in the template, print it elsewhere with \subjectname
\keywords{} % Keywords for your thesis, this is not currently used anywhere in the template, print it elsewhere with \keywordnames
\university{\href{https://www.lshtm.ac.uk/}{London School of Hygiene and Tropical Medicine}} % Your university's name and URL, this is used in the title page and abstract, print it elsewhere with \univname
\department{\href{}{Page count: 43 from Introduction to Conclusions}} % Your department's name and URL, this is used in the title page and abstract, print it elsewhere with \deptname
\group{Candidate number: 110765} % Your research group's name and URL, this is used in the title page, print it elsewhere with \groupname
\faculty{\href{http://faculty.university.com}{Faculty Name}} % Your faculty's name and URL, this is used in the title page and abstract, print it elsewhere with \facname

@@

\begin{declaration}
\addchaptertocentry{\authorshipname} % Add the declaration to the table of contents
\noindent I, declare that this thesis titled, \enquote{\ttitle} and the work presented in it are my own. I confirm that:

@@

%\noindent Signed: \;\;\;\;\;\;\includegraphics[height=2.5\baselineskip]{Figures/signature.png}\\
%\rule[0.5em]{25em}{0.5pt} % This prints a line for the signature
%
%\noindent Date: \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;2018-09-05\\
%\rule[0.5em]{25em}{0.5pt} % This prints a line to write the date
\end{declaration}

@@

\begin{acknowledgements}
\addchaptertocentry{\acknowledgementname} % Add the acknowledgements to the table of contents
\vspace{0.8cm}
I would like to thank my tutor and supervisor \supname, for his guidance, patience, and help while working on this project and also to Dr. Suzana Almoosawi for her invaluable nutritional academic insight. \\

@@

\begin{abstract}
\addchaptertocentry{\abstractname} % Add the abstract to the table of contents
\textbf{Background:}
Recent evidence suggested that there are 3 types of eaters (grazers, early eaters, and late eaters) in the general population according to the timing of their energy consumption occurred throughout the day. This project aimed at investigating the relationship between the timing of eating and carbohydrate intake. Additionally, associations between carbohydrate eating patterns defined previously and hypertension, obesity were explored.

\textbf{Methods:}
The National Diet and Nutrition Survey (NDNS) Rolling Programme (2008/09-15/16) database of detailed four-day food diaries was used in exploration of both timing and quantity eating patterns of carbohydrates in the general adults living in the UK. 6155 participants aged 19 or older were kept in the analysis. Multilevel latent class analysis (MLCA) models were applied in finding the latent classes of carbohydrate consumption accounting for the hierarchical data structure that the observed daily food diaries nested within individuals. Survey-designed multivariable regression models assessed the associations of carbohydrate eating patterns with hypertension, body mass index (BMI), and waist circumferences.

\textbf{Results:}
Three types of carbohydrate eating days (low/high percentage carbohydrate days, and regular meal days) were found in the elementary level, depending on which, three types of carbohydrate eaters (low/moderate/high carbohydrate eaters) emerged in the NDNS RP participants.

\textbf{Conclusion:}

\end{abstract}
